---
title: "Task 7"
output:
  html_document:
    df_print: paged
  pdf_document: default
editor_options: 
  markdown: 
    wrap: sentence
---

# Prerequisites

## Import libraries

```{r}
library(dplyr)
library(ggplot2)
library(car)
library(rstatix)
```

## Import datasets

```{r}
urlP <- "https://raw.githubusercontent.com/WHPAN0108/BHT-DataScience-S24/main/data_clean/data/patient_info.csv"
urlH <- "https://raw.githubusercontent.com/WHPAN0108/BHT-DataScience-S24/main/data_clean/data/date.csv"

patients <- read.csv(urlP)
hospital <- read.csv(urlH)

head(patients)
head(hospital)
```

# Data Clean & EDA

## Task 1

### Replace missing BMI values with the median BMI for the corresponding gender.

Calculate the BMI for male and female seperately.
Exclude na-values.

```{r}
BMImaleMedian <- median( (subset(patients, Sex %in% c('Male') & !is.na(BMI))$BMI ) )
BMIfemaleMedian <- median( (subset(patients, Sex %in% c('Female') & !is.na(BMI))$BMI ) )

print(sprintf("Male median BMI is: %f and female median BMI is: %g", BMImaleMedian, BMIfemaleMedian))
```

Replace BMI values that are na and check afterwards, if any na values for BMI are left.

```{r}
patients$BMI[patients$Sex %in% c('Male') & is.na(patients$BMI)] <- BMImaleMedian
patients$BMI[patients$Sex %in% c('Female') & is.na(patients$BMI)] <- BMIfemaleMedian
subset(patients, is.na(BMI))

```

## Task 2

### Perform a sanity check on all columns in both files, and remove any samples that do not pass.

#### patients

Do a summary and check for unusual or unrealistic values.

```{r}
summary(patients)

```

Anomalies: The minimum age is 6, but the dataset should only contain patients that are 40 or older.
All BMI values should be between 15-60

------------------------------------------------------------------------

Checking for further na-values.
These can be found for "Smoking_status".

```{r}
patients %>% summarize_all(~ sum(is.na(.)))


```

------------------------------------------------------------------------

Check distinct values for colums "Sex", "Smoking_status","Disease".
There are only the expected values.

```{r}
patients %>% distinct(Sex)
patients %>% distinct(Smoking_status)
patients %>% distinct(Disease)
```

#### hospital visits

Previous to checking transform the VisitDate to date-format.

```{r}
hospital <- hospital %>% mutate(v_date = as.Date(VisitDate, "%d/%m/%Y"))
summary(hospital)
hospital %>% summarize_all(~ sum(is.na(.)))

```

Anomalies: There is at least one visit date before 1950 and one after the end of 2023.

------------------------------------------------------------------------

data clean: hospital visits

```{r}

# Remove lines with visits before 1950 or after the end of 2023 or with na-values.
hospital <- hospital %>% filter(!v_date < "1950/01/01", !v_date > "2023/12/31", !is.na(v_date))

```

------------------------------------------------------------------------

data clean: patients

```{r}
#Remove lines where age is <40 or BMI not between 15 to 60 or Smoking_Status is na
patients <- patients %>% filter(!Age < 40 , !BMI < 15 , !BMI > 60, !is.na(Smoking_status))

```

## Task 3

### Calculate the number of visits each patient made since 1/1/2020. If the VisitDate is missing, it indicates no visits during this period.

```{r}
hospitalvisits <- hospital %>% select(Participant_ID, v_date) %>% filter(v_date >= "2020/1/1") %>% group_by(Participant_ID) %>% count(Participant_ID, sort = TRUE)

hospitalvisits <- rename(hospitalvisits, "NumberOfVisits" = n)

# Show results table
hospitalvisits %>% filter(TRUE)
```

## Task 4

### Create a summary table and/or chart showing the number of visits for each month (irrespective of the year).

```{r}
# Add a new colums with only the month of the visitDate
hospital <- hospital %>% mutate(month = format(v_date, "%m") )

# Result as a table
hospital %>% select(month) %>% group_by(month) %>% count(month)

# Save result separately
VisitsPerMonth <- hospital %>% select(month) %>% group_by(month) %>% count(month)

# Draw a bar diagram
ggplot(data=VisitsPerMonth, aes(x=month, y=n)) + geom_bar(stat="identity")

```

## Task 5

### add a new column "age group " to split the age to 5 categories: 40-49, 50-59, 60-69, and above 70)

```{r}
patients <- patients %>% mutate(AgeGroup = case_when(
                                           (Age >= 40 & Age <= 49) ~ "40-49"
                                         , (Age >= 50 & Age <= 59) ~ "50-59"
                                         , (Age >= 60 & Age <= 69) ~ "60-69"
                                         , (Age >= 70) ~ "above 70"
                                         , TRUE ~ NA) )

# Show result table
patients %>% filter(TRUE)



```

## Task 6

### Merge the two files to include the columns: Patient_ID, Sex, Age, BMI, Smoking_status, Number_of_visits, Disease and age group

```{r}
patients_overall =
  patients %>% 
  left_join(hospitalvisits,
    by = join_by(Participant_ID),
    copy = FALSE,
    keep = NULL
  ) %>% 
  mutate(NumberOfVisits = case_when(is.na(NumberOfVisits) ~ 0, TRUE ~ NumberOfVisits)) 
```

# Statistics

## Task 1

### Compare the BMI between the healthy and control groups using a box or violin plot.

```{r}
# Convert Disease from numeric to a factor
#  factor 1 means disease, 2 means healthy
patients$Disease <- as.factor(patients$Disease)

# Draw a boxplot
ggplot(data=patients,aes(x=Disease, y=BMI, color=Disease)) + geom_boxplot() +theme_minimal()

```

Conclusion: The BMI results between patients with and without a disease look mostly the same.
There is no obvious difference.
Median, upper and lower quartile etc. are nearly the same.

## Task 2

### Examine if the proportion of smoking status differs between the healthy and control groups.

Chi-square test

```{r}
#  factor 1 means disease, 2 means healthy
patients_chi <- table(patients$Disease, patients$Smoking_status)
#Table of Disease and smoking status:
print(patients_chi)
chi_squared_test <- chisq.test(patients_chi)

#Chi-Squared Test Results:
print(chi_squared_test)
```

Conclusion: The p-value is not less than 0.05, which indicates, that there is NO significant difference in smoking status between the compared groups.

## Task 3

### Analyze the relationship between smoking status and the number of hospital visits, separately for the disease and healthy groups, and for the overall population.

ANOVA Test

```{r}
#  factor 1 means disease, 2 means healthy

#H0: There is no dependency between the smoking status and the number of visits
#Ha: There is an significant influence of the smoking status on the number of hospital visits.

# ANOVA for each group
anova_healthy <- aov(NumberOfVisits ~ Smoking_status, data = patients_overall %>% filter(Disease == 2))
anova_disease <- aov(NumberOfVisits ~ Smoking_status, data = patients_overall %>% filter(Disease == 1))
anova_overall <- aov(NumberOfVisits ~ Smoking_status, data = patients_overall )

#ANOVA Results per Group:
summary(anova_healthy)
summary(anova_disease)
summary(anova_overall)



```

The p-value (Pr(\>F)) of all three tests differs quite a lot, but none of the p-values is below 0.05.
Therefor there is no significant dependency between the smoking status and the number of hospital visits.
You can reject Ha.

## Task 4

### Determine if the number of visits is higher in the disease group compared to the healthy group.

H0: The number of visits in the disease group is not higher than in the healthy group

Ha: The number of visits in the disease group is higher/greater than in the healthy group

```{r}

patients_overall %>% select(Disease, NumberOfVisits) %>% group_by(Disease) %>% get_summary_stats(NumberOfVisits, type = "mean_sd")

boxplot(NumberOfVisits~Disease, data=patients_overall) 

t_test <- t.test(NumberOfVisits ~ Disease, data = patients_overall, alternative = "greater" )


print(t_test)


```

Conclusion: The summary, as well as the boxplot show pretty much no difference in mean-values.
The p-value is not below 0.05.
You can reject Ha.

## Task 5

### Calculate the disease prevalence (number of sick / total number population) across different age groups in our dataset, and test if the prevalence is different across the group

```{r}
#  factor 1 means disease, 2 means healthy

# get the number of people with disease (disease == 1) per AgeGroup
DiseasePerAgeGroup <- patients_overall %>% group_by(AgeGroup) %>% filter(Disease == 1) %>% count(Disease)

# calculate the prevelance per AgeGroup
DiseasePrevelance <-
  patients_overall %>% 
    group_by(AgeGroup) %>% 
    summarise(Total = n()) %>% 
    left_join(DiseasePerAgeGroup,
              by = join_by(AgeGroup),
              copy = FALSE,
              keep = NULL
              ) %>% 
    select(AgeGroup, Total, n) %>%
    mutate(prevelance = n/Total)


#prevelance_chi <- table(DiseasePrevelance$AgeGroup, DiseasePrevelance$prevelance)
#Table of AgeGroup and prevelance:
#print(prevelance_chi)
#chi_squared_test_prevelance <- chisq.test(prevelance_chi)

#Chi-Squared Test Results:
#print(chi_squared_test_prevelance)

disease_chi <- table(patients$AgeGroup, patients$Disease)
#Table of AgeGroup and DiseaseStatus:
print(disease_chi)
chi_squared_test_disease <- chisq.test(disease_chi)

#Chi-Squared Test Results:
print(chi_squared_test_disease)

```

Conclusion: The p-value is above 0.05 so there is no significant relationship between disease status and AgeGroup.
